---
title: "E-Commerce Data Analysis — Capstone Project"
author: "KIBIDI"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    theme: flatly
    highlight: tango
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo       = TRUE,
  message    = FALSE,
  warning    = FALSE,
  fig.width  = 9,
  fig.height = 5.5,
  fig.align  = "center"
)
```

---

# 1. Introduction

This project applies data science techniques to an e-commerce dataset of **1,000 customer transactions**. The analysis covers customer demographics, purchasing behavior, product performance, payment trends, and predictive modelling for sales and order outcomes.

---


```{r}
setwd("C:/Users/h6p/Desktop/AI FOR DATA ANALYSIS")
```


# 2. Load Libraries & Data

```{r libraries}
library(tidyverse)
library(lubridate)
library(scales)
library(patchwork)
library(corrplot)
library(randomForest)
library(caret)
library(knitr)
library(kableExtra)

```


 #Load data
```{r}
ecommerce <- read_csv("ecommerce_data.csv", show_col_types = FALSE)

glimpse(ecommerce)
```
```{r}
head(ecommerce)
```

```{r}

ecommerce |>
  summarise(across(everything(), ~sum(is.na(.)))) |>
  tidyr::pivot_longer(everything(), names_to = "column", values_to = "missing") |>
  mutate(
    total = nrow(ecommerce),
    pct_missing = round(missing / total * 100, 1)
  ) |>
  filter(missing > 0) |>
  arrange(desc(missing))
```

# 3. Data Cleaning & Preprocessing
```{r cleaning}
ecommerce <- ecommerce %>%
  mutate(
    Order_Date       = ymd(Order_Date),
    Order_Month      = month(Order_Date, label = TRUE),
    Order_Quarter    = quarter(Order_Date),
    Order_Year       = year(Order_Date),
    Age_Group        = cut(Age,
                           breaks = c(17, 25, 35, 45, 55, 70),
                           labels = c("18–25", "26–35", "36–45", "46–55", "56–70")),
    Discount_Applied = Discount > 0,
    Revenue          = Purchase_Amount,
    across(where(is.character), as.factor)
  )

```




```{r}
# Missing values summary
ecommerce %>%
  summarise(across(everything(), ~sum(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "Variable", values_to = "Missing") %>%
  filter(Missing > 0) %>%
  kable(caption = "Missing Values per Column") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```



> **No missing values detected.** The dataset is clean and ready for analysis.

---

# 4. Exploratory Data Analysis (EDA)

## 4.1 Dataset Overview

```{r overview-table}
ecommerce %>%
  summarise(
    Customers        = n_distinct(Customer_ID),
    Total_Orders     = n(),
    Total_Revenue    = dollar(sum(Revenue)),
    Avg_Order_Value  = dollar(mean(Revenue)),
    Avg_Rating       = round(mean(Customer_Rating), 2),
    Date_Range       = paste(min(Order_Date), "→", max(Order_Date))
  ) %>%
  pivot_longer(everything(), names_to = "Metric", values_to = "Value") %>%
  kable(caption = "Key Business Metrics") %>%
  kable_styling(bootstrap_options = c("striped", "condensed"), full_width = FALSE)
```

## 4.2 Customer Demographics

```{r demographics, fig.height=5}
p1 <- ecommerce %>%
  count(Gender) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = "", y = pct, fill = Gender)) +
  geom_col(width = 1, color = "white", linewidth = 0.8) +
  coord_polar("y") +
  geom_text(aes(label = percent(pct, 1)), position = position_stack(vjust = 0.5),
            size = 5, fontface = "bold", color = "white") +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  labs(title = "Gender Distribution", fill = NULL) +
  theme_void(base_size = 13) +
  theme(legend.position = "bottom", plot.title = element_text(hjust = 0.5, face = "bold"))

p2 <- ecommerce %>%
  count(Age_Group) %>%
  ggplot(aes(x = Age_Group, y = n, fill = Age_Group)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = n), vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_brewer(palette = "Set2") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Customers by Age Group", x = "Age Group", y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p1 + p2
```

## 4.3 Purchase Amount Distribution

```{r purchase-dist}
p3 <- ggplot(ecommerce, aes(x = Purchase_Amount)) +
  geom_histogram(aes(y = after_stat(density)), bins = 35,
                 fill = "#4E79A7", color = "white", alpha = 0.85) +
  geom_density(color = "#E15759", linewidth = 1.1) +
  geom_vline(xintercept = mean(ecommerce$Purchase_Amount),
             linetype = "dashed", color = "#F28E2B", linewidth = 1) +
  annotate("text", x = mean(ecommerce$Purchase_Amount) + 80,
           y = 0.00115, label = paste0("Mean: ", dollar(mean(ecommerce$Purchase_Amount))),
           color = "#F28E2B", size = 4, fontface = "bold") +
  labs(title = "Distribution of Purchase Amount",
       x = "Purchase Amount (USD)", y = "Density") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p4 <- ggplot(ecommerce, aes(x = Gender, y = Purchase_Amount, fill = Gender)) +
  geom_violin(alpha = 0.7, trim = FALSE) +
  geom_boxplot(width = 0.15, fill = "white", color = "grey40") +
  scale_fill_manual(values = c("#4E79A7", "#F28E2B")) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Purchase Amount by Gender",
       x = NULL, y = "Purchase Amount (USD)") +
  theme_minimal(base_size = 13) +
  theme(legend.position = "none", plot.title = element_text(face = "bold"))

p3 + p4
```

---

# 5. Product Category Analysis

```{r category-analysis, fig.height=6}
cat_summary <- ecommerce %>%
  group_by(Product_Category) %>%
  summarise(
    Orders    = n(),
    Revenue   = sum(Purchase_Amount),
    Avg_Price = mean(Product_Price),
    Avg_Rating = mean(Customer_Rating),
    .groups = "drop"
  ) %>%
  arrange(desc(Revenue))

p5 <- cat_summary %>%
  ggplot(aes(x = fct_reorder(Product_Category, Revenue), y = Revenue, fill = Product_Category)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = dollar(Revenue, scale = 1e-3, suffix = "K")),
            hjust = -0.1, size = 3.8, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.18))) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Total Revenue by Product Category",
       x = NULL, y = "Revenue (USD)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p6 <- cat_summary %>%
  ggplot(aes(x = fct_reorder(Product_Category, Avg_Rating), y = Avg_Rating, fill = Product_Category)) +
  geom_col(show.legend = FALSE, width = 0.7) +
  geom_text(aes(label = round(Avg_Rating, 2)), hjust = -0.15, size = 3.8, fontface = "bold") +
  coord_flip() +
  scale_y_continuous(limits = c(0, 4), expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Avg Customer Rating by Category",
       x = NULL, y = "Average Rating (1–5)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p5 + p6
```

---

# 6. Sales Trend Analysis

## 6.1 Monthly Revenue Trend

```{r monthly-trend, fig.height=5}
monthly <- ecommerce %>%
  group_by(Order_Month) %>%
  summarise(Revenue = sum(Purchase_Amount), Orders = n(), .groups = "drop")

ggplot(monthly, aes(x = Order_Month, y = Revenue, group = 1)) +
  geom_area(fill = "#4E79A7", alpha = 0.25) +
  geom_line(color = "#4E79A7", linewidth = 1.3) +
  geom_point(aes(size = Orders), color = "#E15759", show.legend = TRUE) +
  scale_y_continuous(labels = dollar) +
  scale_size_continuous(range = c(3, 8), name = "# Orders") +
  labs(title = "Monthly Revenue & Order Volume",
       x = "Month", y = "Total Revenue (USD)") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")
```

## 6.2 Revenue by Category Over Time

```{r heatmap, fig.height=5}
ecommerce %>%
  group_by(Order_Month, Product_Category) %>%
  summarise(Revenue = sum(Purchase_Amount), .groups = "drop") %>%
  ggplot(aes(x = Order_Month, y = Product_Category, fill = Revenue)) +
  geom_tile(color = "white", linewidth = 0.5) +
  scale_fill_gradient(low = "#EFF3FF", high = "#084594", labels = dollar) +
  labs(title = "Revenue Heatmap: Category × Month",
       x = "Month", y = NULL, fill = "Revenue") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        axis.text.x = element_text(angle = 45, hjust = 1))
```

---

# 7. Payment & Order Status Analysis

```{r payment-status, fig.height=5}
p7 <- ecommerce %>%
  count(Payment_Method) %>%
  ggplot(aes(x = fct_reorder(Payment_Method, n), y = n, fill = Payment_Method)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = n), hjust = -0.2, size = 4, fontface = "bold") +
  coord_flip() +
  scale_fill_brewer(palette = "Paired") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Orders by Payment Method", x = NULL, y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p8 <- ecommerce %>%
  count(Order_Status) %>%
  mutate(pct = n / sum(n)) %>%
  ggplot(aes(x = fct_reorder(Order_Status, pct), y = pct, fill = Order_Status)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = percent(pct, 1)), hjust = -0.2, size = 4, fontface = "bold") +
  coord_flip() +
  scale_fill_manual(values = c("#59A14F", "#E15759", "#F28E2B", "#76B7B2")) +
  scale_y_continuous(labels = percent, expand = expansion(mult = c(0, 0.15))) +
  labs(title = "Order Status Breakdown", x = NULL, y = "Proportion") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p7 + p8
```

---

# 8. Discount & Pricing Impact

```{r discount-impact, fig.height=5}
p9 <- ecommerce %>%
  mutate(Discount_Band = cut(Discount, breaks = c(-1, 0, 10, 20, 30, 50),
                              labels = c("None", "1–10%", "11–20%", "21–30%", "31–50%"))) %>%
  group_by(Discount_Band) %>%
  summarise(Avg_Revenue = mean(Purchase_Amount), .groups = "drop") %>%
  ggplot(aes(x = Discount_Band, y = Avg_Revenue, fill = Discount_Band)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  geom_text(aes(label = dollar(Avg_Revenue)), vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_brewer(palette = "Blues", direction = 1) +
  scale_y_continuous(labels = dollar, expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Avg Revenue by Discount Band",
       x = "Discount Range", y = "Avg Purchase Amount") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p10 <- ggplot(ecommerce, aes(x = Product_Price, y = Purchase_Amount, color = Product_Category)) +
  geom_point(alpha = 0.4, size = 1.8) +
  geom_smooth(method = "lm", se = FALSE, linewidth = 0.9, color = "black") +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  scale_color_brewer(palette = "Set1") +
  labs(title = "Product Price vs Purchase Amount",
       x = "Product Price", y = "Purchase Amount", color = "Category") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"),
        legend.position = "right")

p9 + p10
```

---

# 9. Customer Spending & Segmentation

```{r segmentation, fig.height=5}
customer_seg <- ecommerce %>%
  group_by(Customer_ID) %>%
  summarise(
    Total_Spend   = sum(Purchase_Amount),
    Num_Orders    = n(),
    Avg_Rating    = mean(Customer_Rating),
    .groups = "drop"
  ) %>%
  mutate(Segment = case_when(
    Total_Spend >= quantile(Total_Spend, 0.75) ~ "High Value",
    Total_Spend >= quantile(Total_Spend, 0.25) ~ "Mid Value",
    TRUE                                        ~ "Low Value"
  ))

p11 <- customer_seg %>%
  count(Segment) %>%
  ggplot(aes(x = Segment, y = n, fill = Segment)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = n), vjust = -0.4, size = 4.5, fontface = "bold") +
  scale_fill_manual(values = c("High Value" = "#59A14F", "Mid Value" = "#F28E2B", "Low Value" = "#E15759")) +
  scale_y_continuous(expand = expansion(mult = c(0, 0.12))) +
  labs(title = "Customer Segments by Spend", x = NULL, y = "Count") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p12 <- ggplot(customer_seg, aes(x = Num_Orders, y = Total_Spend, color = Segment)) +
  geom_point(alpha = 0.6, size = 2.5) +
  scale_y_continuous(labels = dollar) +
  scale_color_manual(values = c("High Value" = "#59A14F", "Mid Value" = "#F28E2B", "Low Value" = "#E15759")) +
  labs(title = "Orders vs Total Spend (by Segment)",
       x = "Number of Orders", y = "Total Spend", color = "Segment") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p11 + p12
```

---

# 10. Correlation Analysis

```{r correlation, fig.height=5.5}
numeric_vars <- ecommerce %>%
  select(Age, Product_Price, Quantity, Discount, Shipping_Cost, Purchase_Amount, Customer_Rating)

cor_matrix <- cor(numeric_vars, use = "complete.obs")

corrplot(cor_matrix,
         method     = "color",
         type       = "upper",
         order      = "hclust",
         tl.col     = "black",
         tl.srt     = 45,
         addCoef.col = "black",
         number.cex = 0.75,
         col        = colorRampPalette(c("#E15759", "white", "#4E79A7"))(200),
         title      = "Correlation Matrix — Numeric Variables",
         mar        = c(0, 0, 2, 0))
```

---

# 11. Machine Learning

## 11.1 Feature Engineering

```{r feature-engineering}
ecommerce_ml <- ecommerce %>%
  mutate(
    High_Value    = ifelse(Purchase_Amount > median(Purchase_Amount), 1, 0),
    Completed     = ifelse(Order_Status == "Completed", 1, 0),
    Gender_Num    = ifelse(Gender == "Male", 1, 0),
    Disc_Applied  = ifelse(Discount > 0, 1, 0)
  ) %>%
  select(Age, Product_Price, Quantity, Discount, Shipping_Cost,
         Customer_Rating, Gender_Num, Disc_Applied,
         High_Value, Completed, Purchase_Amount) %>%
  na.omit()
```

## 11.2 Model 1 — Purchase Amount Prediction (Regression)

```{r regression}
set.seed(42)
idx <- createDataPartition(ecommerce_ml$Purchase_Amount, p = 0.8, list = FALSE)
train_r <- ecommerce_ml[idx, ]
test_r  <- ecommerce_ml[-idx, ]

rf_reg <- randomForest(Purchase_Amount ~ Age + Product_Price + Quantity +
                         Discount + Shipping_Cost + Customer_Rating +
                         Gender_Num + Disc_Applied,
                       data  = train_r,
                       ntree = 300,
                       importance = TRUE)

pred_r <- predict(rf_reg, test_r)

reg_metrics <- data.frame(
  Metric = c("MAE", "RMSE", "R²"),
  Value  = c(
    round(mean(abs(pred_r - test_r$Purchase_Amount)), 2),
    round(sqrt(mean((pred_r - test_r$Purchase_Amount)^2)), 2),
    round(1 - sum((pred_r - test_r$Purchase_Amount)^2) /
            sum((test_r$Purchase_Amount - mean(test_r$Purchase_Amount))^2), 4)
  )
)

kable(reg_metrics, caption = "Regression Model Performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r reg-plots, fig.height=5}
reg_df <- data.frame(Actual = test_r$Purchase_Amount, Predicted = pred_r)

p_reg1 <- ggplot(reg_df, aes(x = Actual, y = Predicted)) +
  geom_point(alpha = 0.5, color = "#4E79A7", size = 2) +
  geom_abline(color = "#E15759", linewidth = 1.1, linetype = "dashed") +
  scale_x_continuous(labels = dollar) +
  scale_y_continuous(labels = dollar) +
  labs(title = "Actual vs Predicted — Purchase Amount",
       x = "Actual", y = "Predicted") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

imp_df <- as.data.frame(importance(rf_reg)) %>%
  rownames_to_column("Feature") %>%
  arrange(desc(`%IncMSE`))

p_reg2 <- ggplot(imp_df, aes(x = fct_reorder(Feature, `%IncMSE`), y = `%IncMSE`, fill = `%IncMSE`)) +
  geom_col(show.legend = FALSE, width = 0.65) +
  coord_flip() +
  scale_fill_gradient(low = "#AEC6E8", high = "#4E79A7") +
  labs(title = "Feature Importance (Regression)",
       x = NULL, y = "% Increase in MSE") +
  theme_minimal(base_size = 13) +
  theme(plot.title = element_text(face = "bold"))

p_reg1 + p_reg2
```

## 11.3 Model 2 — High-Value Purchase Classification

```{r classification}
set.seed(42)
ecommerce_ml$High_Value <- as.factor(ecommerce_ml$High_Value)

idx2   <- createDataPartition(ecommerce_ml$High_Value, p = 0.8, list = FALSE)
train_c <- ecommerce_ml[idx2, ]
test_c  <- ecommerce_ml[-idx2, ]

rf_clf <- randomForest(High_Value ~ Age + Product_Price + Quantity +
                         Discount + Shipping_Cost + Customer_Rating +
                         Gender_Num + Disc_Applied,
                       data  = train_c,
                       ntree = 300)

pred_c <- predict(rf_clf, test_c)
cm     <- confusionMatrix(pred_c, test_c$High_Value)

clf_metrics <- data.frame(
  Metric    = c("Accuracy", "Precision", "Recall", "F1-Score"),
  Value     = round(c(
    cm$overall["Accuracy"],
    cm$byClass["Precision"],
    cm$byClass["Recall"],
    cm$byClass["F1"]
  ), 4)
)

kable(clf_metrics, caption = "Classification Model Performance") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE)
```

```{r conf-matrix-plot, fig.height=5}
cm_df <- as.data.frame(cm$table) %>%
  rename(Predicted = Prediction, Actual = Reference)

ggplot(cm_df, aes(x = Actual, y = Predicted, fill = Freq)) +
  geom_tile(color = "white", linewidth = 1) +
  geom_text(aes(label = Freq), size = 7, fontface = "bold", color = "white") +
  scale_fill_gradient(low = "#AEC6E8", high = "#084594") +
  scale_x_discrete(labels = c("0" = "Low Value", "1" = "High Value")) +
  scale_y_discrete(labels = c("0" = "Low Value", "1" = "High Value")) +
  labs(title = "Confusion Matrix — High-Value Classifier",
       fill = "Count") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(face = "bold", hjust = 0.5))
```

---

# 12. Key Insights & Recommendations

```{r insights-table, echo=FALSE}
insights <- data.frame(
  `#`          = 1:7,
  Insight      = c(
    "Electronics and Home generate the highest total revenue.",
    "Customer ratings are evenly spread (avg ≈ 3.0) — no strong satisfaction skew.",
    "M-Pesa and Credit Card dominate payments, reflecting local & global preferences.",
    "Higher discount bands do not consistently boost average purchase amount.",
    "Product price is the strongest predictor of purchase amount in ML models.",
    "~25% of orders are returned or canceled — a key area for operational improvement.",
    "Mid-age customers (36–55) form the largest buying group."
  ),
  Recommendation = c(
    "Prioritise stock and promotions for Electronics and Home categories.",
    "Introduce loyalty programs to push ratings above 4 consistently.",
    "Optimize checkout for M-Pesa and Credit Card; remove friction elsewhere.",
    "Re-evaluate discount strategy — focus on targeted rather than blanket discounts.",
    "Use price-tier promotions to shift customers to higher-value purchases.",
    "Audit fulfillment processes to reduce returns and cancellations.",
    "Target 36–55 age bracket with personalized campaigns."
  )
)

kable(insights, col.names = c("#", "Insight", "Recommendation"),
      caption = "Key Business Insights & Recommendations") %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"), full_width = TRUE) %>%
  column_spec(2, italic = TRUE) %>%
  column_spec(3, bold = TRUE)
```

---

# 13. Conclusion

This capstone project demonstrated a complete data science pipeline applied to e-commerce data:

- **Data Cleaning** — the dataset was clean with no missing values; date and categorical variables were processed appropriately.
- **EDA** — revealed key trends in customer demographics, product performance, payment behavior, and seasonal sales.
- **Machine Learning** — Random Forest models achieved strong performance for both revenue regression and high-value purchase classification.
- **Business Value** — actionable insights were generated around discounting, category strategy, payment optimization, and customer retention.

The project confirms that data-driven decision-making can meaningfully improve e-commerce operations, customer experience, and revenue growth.

---

*Report generated with R `r R.version$major`.`r R.version$minor` | `r Sys.Date()`*
