---
title: "Capstone Project 2: AI : E-Commerce"
author: "KIBIDI"
date: "2026-05-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown




1. Data Collection and Loading

#1.0:Set Working Directory 
```{r}
setwd("C:/Users/h6p/Desktop/ALCHYMED DATA ANALYSIS/AI")
```


#1.1:Load Libraries
```{r}
library(tidyverse)
library(ggplot2)
library(dplyr)
library(lubridate)
library(skimr)
library(corrplot)
library(caret)
library(randomForest)
library(forecast)
library(janitor)

theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(size = 12),
      axis.title = element_text(face = "bold"),
      legend.title = element_text(face = "bold")
    )
)
```


#1.2: Load Dataset
```{r load-data}
ecommerce_data <- read.csv("ecommerce_data.csv", stringsAsFactors = FALSE)
head(ecommerce_data)
```


#1.3a: Explore Data :Structure
```{r}
str(ecommerce_data)
```


#1.3b: Explore Data :Statistics
```{r}
skimr::skim(ecommerce_data)
```


2. Data Cleaning and Preprocessing
#2.1:missing values
```{r}
ecommerce_clean <- ecommerce_data %>%
  janitor::clean_names() %>%
  distinct() %>%
  mutate(
    order_date = mdy(order_date),
    gender = as.factor(gender),
    location = as.factor(location),
    product_category = as.factor(product_category),
    payment_method = as.factor(payment_method),
    order_status = as.factor(order_status)
  )

num_cols <- ecommerce_clean %>% select(where(is.numeric)) %>% names()
cat_cols <- ecommerce_clean %>% select(where(~ is.character(.x) || is.factor(.x))) %>% names()

for (col_name in num_cols) {
  ecommerce_clean[[col_name]][is.na(ecommerce_clean[[col_name]])] <- median(ecommerce_clean[[col_name]], na.rm = TRUE)
}

for (col_name in cat_cols) {
  mode_val <- names(sort(table(ecommerce_clean[[col_name]]), decreasing = TRUE))[1]
  ecommerce_clean[[col_name]][is.na(ecommerce_clean[[col_name]])] <- mode_val
}

sum(is.na(ecommerce_clean))
```

#3. Feature Engineering 
```{r}
ecommerce_clean <- ecommerce_clean %>%
  mutate(
    order_year = year(order_date),
    order_month = month(order_date, label = TRUE, abbr = TRUE),
    order_quarter = quarter(order_date),
    day_of_week = wday(order_date, label = TRUE, abbr = TRUE),
    customer_segment = case_when(
      age < 30 ~ "Young",
      age < 50 ~ "Adult",
      TRUE ~ "Senior"
    ),
    discount_rate = if_else(product_price > 0, discount / product_price, 0),
    high_value_order = if_else(purchase_amount >= median(purchase_amount, na.rm = TRUE), "Yes", "No"),
    net_revenue_est = purchase_amount - shipping_cost,
    quantity_band = case_when(
      quantity == 1 ~ "Single",
      quantity <= 3 ~ "Moderate",
      TRUE ~ "Bulk"
    )
  ) %>%
  mutate(
    customer_segment = as.factor(customer_segment),
    high_value_order = as.factor(high_value_order),
    quantity_band = as.factor(quantity_band)
  )

head(ecommerce_clean)
```


#summary statistics 
```{r}
summary(ecommerce_clean %>% select(age, product_price, quantity, discount, shipping_cost, purchase_amount, customer_rating, net_revenue_est))
```


## Missing Values and Duplicate Check
```{r}
quality_summary <- tibble(
  metric = c("Rows after cleaning", "Columns", "Remaining missing values", "Duplicate rows after cleaning"),
  value = c(nrow(ecommerce_clean), ncol(ecommerce_clean), sum(is.na(ecommerce_clean)), sum(duplicated(ecommerce_clean)))
)
quality_summary
```

```{r}
str(ecommerce_clean)
```



3.Exploratory Data Analysis

# 3.1a:Age DistributioN
```{r}
ggplot(ecommerce_clean, aes(x = age)) +
  geom_histogram(fill = "brown", color = "white", bins = 25) +
  labs(title = "Customer Age Distribution", x = "Age", y = "Count")
```

# 3.1b:Purchase Amount Distribution
```{r}
ggplot(ecommerce_clean, aes(x = purchase_amount)) +
  geom_histogram(fill = "darkcyan", color = "white", bins = 30) +
  labs(title = "Distribution of Purchase Amount", x = "Purchase Amount", y = "Frequency")
```


#3.1c:Purchase Amount by Gender boxplot
```{r}
ggplot(ecommerce_clean, aes(x = gender, y = purchase_amount, fill = gender)) +
  geom_boxplot(alpha = 0.8) +
  labs(title = "Purchase Amount by Gender", x = "Gender", y = "Purchase Amount") +
  guides(fill = "none")
```


#3.1d:Customer Rating Distribution by Order Status
```{r}
ggplot(ecommerce_clean, aes(x = order_status, y = customer_rating, fill = order_status)) +
  geom_boxplot(alpha = 0.85) +
  labs(title = "Customer Rating by Order Status", x = "Order Status", y = "Customer Rating") +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

#3.1e:Correlation Analysis
```{r}
correlation_data <- ecommerce_clean %>%
  select(age, product_price, quantity, discount, shipping_cost, purchase_amount, customer_rating, order_quarter, discount_rate, net_revenue_est)

cor_matrix <- cor(correlation_data, use = "complete.obs")

corrplot(cor_matrix, method = "color", type = "upper", addCoef.col = "black", tl.col = "black", tl.srt = 45)
```

4:Customer Behavior Analysis

#4.1:Customer Segment Spending
```{r segment-spending}
segment_summary <- ecommerce_clean %>%
  group_by(customer_segment) %>%
  summarise(
    avg_purchase = mean(purchase_amount),
    avg_quantity = mean(quantity),
    avg_rating = mean(customer_rating),
    orders = n(),
    .groups = "drop"
  )

segment_summary

ggplot(segment_summary, aes(x = customer_segment, y = avg_purchase, fill = customer_segment)) +
  geom_col(width = 0.7) +
  labs(title = "Average Purchase Amount by Customer Segment", x = "Customer Segment", y = "Average Purchase Amount") +
  guides(fill = "none")
```


#4.2:Purchase Frequency Analysis
```{r purchase-frequency}
customer_frequency <- ecommerce_clean %>%
  count(customer_id, name = "purchase_frequency")

freq_summary <- customer_frequency %>%
  summarise(
    min_frequency = min(purchase_frequency),
    avg_frequency = mean(purchase_frequency),
    max_frequency = max(purchase_frequency)
  )

freq_summary

ggplot(customer_frequency, aes(x = purchase_frequency)) +
  geom_histogram(fill = "mediumpurple", color = "white", bins = 20) +
  labs(title = "Customer Purchase Frequency Distribution", x = "Number of Purchases per Customer", y = "Customer Count")
```



#4.3: High-Value Order Share by Customer Segment
```{r high-value-share}
high_value_segment <- ecommerce_clean %>%
  count(customer_segment, high_value_order) %>%
  group_by(customer_segment) %>%
  mutate(share = n / sum(n)) %>%
  ungroup()

ggplot(high_value_segment, aes(x = customer_segment, y = share, fill = high_value_order)) +
  geom_col(position = "fill") +
  scale_y_continuous(labels = scales::percent_format()) +
  labs(title = "High-Value Order Composition by Customer Segment", x = "Customer Segment", y = "Share", fill = "High-Value Order")
```


#4.4:Sales Trend Analysis
```{r sales-trend-monthly}
monthly_sales <- ecommerce_clean %>%
  mutate(month_start = floor_date(order_date, unit = "month")) %>%
  group_by(month_start) %>%
  summarise(
    total_sales = sum(purchase_amount),
    total_orders = n(),
    avg_sales = mean(purchase_amount),
    .groups = "drop"
  )

monthly_sales

ggplot(monthly_sales, aes(x = month_start, y = total_sales)) +
  geom_line(color = "firebrick", linewidth = 1.2) +
  geom_point(color = "firebrick", size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "navy", linetype = "dashed") +
  labs(title = "Monthly Sales Trend", x = "Month", y = "Total Sales")
```



#4.5: Forecast-Oriented Time Series Plot
```{r time-series-forecast}
sales_ts <- ts(monthly_sales$total_sales, frequency = 12)
auto_model <- forecast::auto.arima(sales_ts)
future_sales <- forecast::forecast(auto_model, h = 3)
plot(future_sales, main = "Short-Term Sales Forecast")
```


5. Regional and Category Analysis

#5.1:Top Product Categories by Sales
```{r category-sales}
category_sales <- ecommerce_clean %>%
  group_by(product_category) %>%
  summarise(total_sales = sum(purchase_amount), total_quantity = sum(quantity), .groups = "drop") %>%
  arrange(desc(total_sales))

category_sales

ggplot(category_sales, aes(x = reorder(product_category, total_sales), y = total_sales, fill = product_category)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Total Sales by Product Category", x = "Product Category", y = "Total Sales")
```


#5.2:Category Sales by Gender
```{r faceted-category-gender}
ggplot(ecommerce_clean, aes(x = product_category, y = purchase_amount, fill = gender)) +
  geom_boxplot(outlier.alpha = 0.3) +
  facet_wrap(~ gender) +
  labs(title = "Purchase Amount by Product Category and Gender", x = "Product Category", y = "Purchase Amount") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
```


#5.3:Top Locations by Sales
```{r location-sales}
location_sales <- ecommerce_clean %>%
  group_by(location) %>%
  summarise(total_sales = sum(purchase_amount), orders = n(), .groups = "drop") %>%
  arrange(desc(total_sales)) %>%
  slice_head(n = 15)

ggplot(location_sales, aes(x = reorder(location, total_sales), y = total_sales, fill = total_sales)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Top 15 Locations by Total Sales", x = "Location", y = "Total Sales")
```



#5.4:Payment Method Analysis
```{r payment-analysis}
payment_summary <- ecommerce_clean %>%
  group_by(payment_method) %>%
  summarise(
    total_sales = sum(purchase_amount),
    avg_purchase = mean(purchase_amount),
    orders = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(total_sales))

payment_summary

ggplot(payment_summary, aes(x = reorder(payment_method, total_sales), y = total_sales, fill = payment_method)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  labs(title = "Total Sales by Payment Method", x = "Payment Method", y = "Total Sales")
```



#5.5:Discount Impact Analysis
```{r discount-impact-scatter}
ggplot(ecommerce_clean, aes(x = discount, y = purchase_amount, color = product_category)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "lm", se = FALSE, color = "black") +
  labs(title = "Discount vs Purchase Amount", x = "Discount", y = "Purchase Amount", color = "Product Category")
```



#5.6:discount-impact-summary
```{r discount-impact-summary}
discount_groups <- ecommerce_clean %>%
  mutate(discount_band = case_when(
    discount == 0 ~ "No Discount",
    discount < quantile(discount[discount > 0], 0.5, na.rm = TRUE) ~ "Lower Discount",
    TRUE ~ "Higher Discount"
  )) %>%
  group_by(discount_band) %>%
  summarise(
    avg_purchase = mean(purchase_amount),
    avg_quantity = mean(quantity),
    orders = n(),
    .groups = "drop"
  )

discount_groups

ggplot(discount_groups, aes(x = discount_band, y = avg_purchase, fill = discount_band)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Average Purchase Amount by Discount Band", x = "Discount Band", y = "Average Purchase Amount")
```


#5.7:Order Status Analysis
```{r order-status-analysis}
status_summary <- ecommerce_clean %>%
  count(order_status) %>%
  mutate(share = n / sum(n))

status_summary

ggplot(status_summary, aes(x = order_status, y = n, fill = order_status)) +
  geom_col(show.legend = FALSE) +
  labs(title = "Order Status Distribution", x = "Order Status", y = "Count")
```

6:Predictive Modeling

#6.1:Regression Model for Sales Prediction
```{r regression-model-data}
regression_data <- ecommerce_clean %>%
  select(purchase_amount, age, gender, product_category, product_price, quantity, discount,
         shipping_cost, customer_rating, payment_method, order_status, order_quarter,
         customer_segment, discount_rate, net_revenue_est) %>%
  drop_na()

set.seed(123)
reg_train_index <- createDataPartition(regression_data$purchase_amount, p = 0.8, list = FALSE)
reg_train <- regression_data[reg_train_index, ]
reg_test <- regression_data[-reg_train_index, ]
```


#6.1:Regression-model-fit
```{r regression-model-fit}
reg_model <- lm(purchase_amount ~ . - net_revenue_est, data = reg_train)
reg_pred <- predict(reg_model, newdata = reg_test)

reg_rmse <- RMSE(reg_pred, reg_test$purchase_amount)
reg_mae <- MAE(reg_pred, reg_test$purchase_amount)
reg_r2 <- R2(reg_pred, reg_test$purchase_amount)

reg_metrics <- tibble(
  Metric = c("RMSE", "MAE", "R-squared"),
  Value = c(reg_rmse, reg_mae, reg_r2)
)
reg_metrics
```

#6.2:Regression-plot
```{r regression-plot}
reg_results <- tibble(actual = reg_test$purchase_amount, predicted = reg_pred)

ggplot(reg_results, aes(x = actual, y = predicted)) +
  geom_point(color = "steelblue", alpha = 0.65) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(title = "Regression Model: Actual vs Predicted Purchase Amount", x = "Actual", y = "Predicted")
```


#6.3:Classification Model for Customer Behavior / Order Outcome
```{r classification-model-data}
classification_data <- ecommerce_clean %>%
  select(order_status, age, gender, product_category, product_price, quantity, discount,
         shipping_cost, purchase_amount, customer_rating, payment_method, order_quarter,
         customer_segment, high_value_order, quantity_band) %>%
  drop_na()

set.seed(123)
class_train_index <- createDataPartition(classification_data$order_status, p = 0.8, list = FALSE)
class_train <- classification_data[class_train_index, ]
class_test <- classification_data[-class_train_index, ]
```


#6.3:classification-model-fit
```{r classification-model-fit}
class_model <- randomForest(order_status ~ ., data = class_train, ntree = 200, importance = TRUE)
class_pred <- predict(class_model, newdata = class_test)
conf_mat <- confusionMatrix(class_pred, class_test$order_status)

accuracy_val <- as.numeric(conf_mat$overall["Accuracy"])
precision_vals <- diag(conf_mat$table) / colSums(conf_mat$table)
recall_vals <- diag(conf_mat$table) / rowSums(conf_mat$table)
f1_vals <- 2 * (precision_vals * recall_vals) / (precision_vals + recall_vals)

class_metrics <- tibble(
  Metric = c("Accuracy", "Precision", "Recall", "F1-score"),
  Value = c(
    accuracy_val,
    mean(precision_vals, na.rm = TRUE),
    mean(recall_vals, na.rm = TRUE),
    mean(f1_vals, na.rm = TRUE)
  )
)
class_metrics
```


#6.3:Classification-matrix-plot
```{r classification-matrix-plot}
conf_df <- as.data.frame(conf_mat$table)

ggplot(conf_df, aes(x = Reference, y = Prediction, fill = Freq)) +
  geom_tile(color = "white") +
  geom_text(aes(label = Freq), color = "black", size = 4) +
  scale_fill_gradient(low = "#dceaf7", high = "#08519c") +
  labs(title = "Confusion Matrix Heatmap", x = "Actual Class", y = "Predicted Class", fill = "Count")
```

#6.4:Variable-importance
```{r variable-importance}
varImpPlot(class_model, main = "Random Forest Variable Importance")
```














